XSGRB - The X-shooter GRB afterglow sample

### 2016-12-31 document/XSGRB_sample.tex
Adding skymap and text

\title{\vspace{-0.5cm} XSGRB - The X-shooter GRB afterglow sample}
\thanks{Based on observations collected at the European Southern Observatory, Paranal, 
	Chile, Program ID: 098.A-0055, 097.A-0036, 096.A-0079, 095.B-0811(B), 
	095.A-0045, 094.A-0134, 093.A-0069, 092.A-0124, 0091.C-0934, 090.A-0088,
	089.A-0067, 088.A-0051, 087.A-0055, 086.A-0073, 085.A-0009 and 084.A-0260}

\titlerunning{XSGRB sample}

\author{
	J.~Selsing\inst{1}, 
	J.~P.~U. Fynbo\inst{1}, 
	D.~Malesani\inst{1}, 
	T.~Kr\"{u}hler\inst{1}, 
	P.~Goldoni\inst{14}, 
	A.~de~Ugarte~Postigo\inst{11}, 
	J.~Japelj\inst{20},
	P.~D'Avanzo,
	Z.~Cano,
	S.~Covino\inst{10}, 
	V.~D'Elia\inst{7, 12}, 
	H.~Flores,
	O.~E.~Hartoog\inst{6},
	J.~Hjorth\inst{1}, 
	P.~Jakobsson\inst{5}, 
	A.~Levan,
	A.~Melandri,
	S.~Piranomonte \inst{7},
	R.~S\'anchez-Ram\'irez\inst{11},
	S.~Schulze\inst{17, 18}, 
	N.~R.~Tanvir\inst{19},
	C.~Th{\"o}ne,
	S.~D.~Vergani\inst{7, 8},
	P.~M.~Vreeswijk\inst{3}, 
	D.~J.~Watson\inst{1},
	K.~Wiersema\inst{19},
	D.~Xu\inst{1}
	L.~Christensen\inst{1},
	A.~De~Cia\inst{3}, 
	L.~Kaper\inst{6}, 
	L.~A.~Antonelli,
	F.~Fiore,
	A.~Gomboc,
	P.~Groot,
	F.~Hammer,
	C.~Ledoux\inst{2}, 
	E.~Maiorano,
	B.~Milvang-Jensen\inst{1}, 
	E.~Palazzi,
	E.~Pian,
	J.~Schaye,
	G.~Tagliaferri\inst{7},
	R.~A.~M.~J.~Wijers\inst{6}
}

@@

%In P84 we started the X-shooter legacy survey in GTO time. Our aim is to  build
%a sample of GRBs superior to our previous  low-resolution  survey, Fynbo et al.\
%2009, ApJS, 185, 526) both in terms of quantity and  quality of the spectra. The
%last period of GTO time on the survey was P90 (plus 6 hr of leftover GTO time
%in  P91) and for P92 a proposal like this was  accepted for continuation of the
%program in open time (concerning P93 see box 8).

@@

X-shooter is in many ways
the ideal GRB follow-up instrument and indeed GRB follow-up was one of the
primary science cases behind the instrument design and implementation. Our
program secures general purpose GRB afterglow spectroscopic follow-up 
that adds strong legacy value to the \textit{Swift} GRB sample.
Due to the wide wavelength range of X-shooter  
with the same observation cover molecular H$_2$ absorption near the
atmospheric cut-off and all the strong emission lines from the host in the
NIR arm (e.g., Friis et al., 2015, MNRAS, 451, 167). In general, the wide wavelength coverage ensures that we always have
features on which to base a redshift measurement as long as the afterglow is
brighter than about 23 mag in either the $R$- or $z$-band. 
Frequently, emission lines are also detected from the underlying host, which
also provide further information such as SFR and metallicity (the top right 
panel in Fig.~1 shows an example). Only for 
7 out of more than 70 secured spectra could we not measure a redshift.   
With the X-shooter survey we
provide {\bf metallicity measurements} for about 30\% (Voigt-profile fits)
of the $z>1.7$ events.
So far we have
measured metallicities for more than 20 GRB afterglows with X-shooter.
With the wide wavelength coverage of X-shooter we can
study important chemical species as Zn\,\textsc{ii}, Cr\,\textsc{ii} and $\alpha$
elements over a much wider redshift range than what is possible with other
instruments.
As an example, we have measured a metallicity of $0.1 Z_\odot$ for GRB\,100219A
at $z=4.669$ (Th{\"o}ne et al.\ 2013, MNRAS, 428, 3590), $0.02Z_\odot$ for 
GRB\,111008A at $z=4.991$ (Sparre et al.\ 2014, ApJ, 785, 150) and $0.05Z_\odot$
for the $z=5.9125$ GRB\,130606A (Hartoog et al.\ 2015, A\&A, 580, 139).  
Reconciling the
abundance patterns of GRB absorbers, other types of absorbers, QSO DLAs
in particular, and old stars in the Local Group is an important long-term
goal (see also Sparre et al.\ 2014, ApJ, 785, 150).
Metallicities are also measured from host emission lines (Kr{\"u}hler et al.\ 2015,
A\&A, 581, A125).
GRB spectroscopy also allows us to determine the dust content of their environments,
both through analysis of the depletion pattern and through measurement of the
associated extinction (Japelj et al.\ 2015, A\&A, 451, 2050). 
This allows us to quantify the dust-to-metals ratio and its
evolution with redshift (e.g., De Cia et al.\ 2013, A\&A, 560, 88; Zafar \&
Watson 2013, A\&A, 560, 26).
%Moreover, GRB
%afterglows allow investigation of random foreground absorbers, complementing
%QSO sightline studies (e.g., Schulze et
%al.\ 2012, A\&A, 546, 20).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Sample selection criteria and observations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Observations} \label{obs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

We list the overview of all the observations in Tab. \ref{tab:sample_overview}.
\input{tables/burst_overview.tex}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sample completeness}	
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Using the observational characteristics of the 1244 \todo{Get number} bursts
observed by \textit{Swift} to date \cite{Gehrels2009}, and the X-ray derived
hydrogen column densities \citep{Evans2009}, we can quantify the degree to which
our sample is biased against the overall \textit{Swift} sample. The values are
queried from then Three main samples are of interest to compare to assess the
completenessof the follow-up campaign; the full \textit{Swift} sample consisting
of all the burst observed by \textit{Swift} (1), the burst that fulfill the
selection criteria imposed in Sect. \ref{samplecrit} (2), and the bursts
actually followed up with X-shooter (3).

@@

Using a
2-sided Kolmogorov-Smirnov (KS), we can assess the degree to which the null that
the two distribution are drawn from the same distribution can be rejected.

@@

@@

How many bursts have been excluded due to particularities in the follow-up attempt. Visitor, Weather, technical issues.

@@

@@

\begin{figure*}
	\centering
	\includegraphics[page=1, width=16cm]{figures/HI_measurements.pdf}
	\caption{ Measurements of the hydrogen column-densities for all bursts with a clear Lyman alpha absorption system. In solid black is shown the spectrum with black dotted giving the corresponding 1-$\sigma$ error. Black dashed shows zero flux density. The solid red line is the absorption of column density equal to the value presented in Tab. \ref{HI_tab} with the 1-$\sigma$ interval shown with dashed lines.}
	\label{fig:HI1}
\end{figure*}

>>>>

\title{XSGRB - The X-shooter GRB afterglow sample}
\thanks{Based on observations collected at the European Southern Observatory, Paranal, 
	Chile, Program ID: 098.A-0055, 097.A-0036, 096.A-0079, 095.B-0811(B), 
	095.A-0045, 094.A-0134, 093.A-0069, 092.A-0124, 0091.C-0934, 090.A-0088,
	089.A-0067, 088.A-0051, 087.A-0055, 086.A-0073, 085.A-0009 and 084.A-0260}

\titlerunning{The X-shooter sample of GRBs}

\author{
	J.~Selsing\inst{1}, 
	D.~Malesani\inst{1}, 
	T.~Kr\"{u}hler\inst{1}, 
	J.~P.~U. Fynbo\inst{1}, 
	P.~Goldoni\inst{14}, 
	A.~de~Ugarte~Postigo\inst{11}, 
	J.~Japelj\inst{20},
	P.~D'Avanzo,
	Z.~Cano,
	S.~Covino\inst{10}, 
	V.~D'Elia\inst{7, 12}, 
	H.~Flores,
	O.~E.~Hartoog\inst{6},
	J.~Hjorth\inst{1}, 
	P.~Jakobsson\inst{5}, 
	A.~Levan,
	A.~Melandri,
	S.~Piranomonte \inst{7},
	R.~S\'anchez-Ram\'irez\inst{11},
	S.~Schulze\inst{17, 18}, 
	N.~R.~Tanvir\inst{19},
	C.~Th{\"o}ne,
	S.~D.~Vergani\inst{7, 8},
	P.~M.~Vreeswijk\inst{3}, 
	D.~J.~Watson\inst{1},
	K.~Wiersema\inst{19},
	D.~Xu\inst{1}
	L.~Christensen\inst{1},
	A.~De~Cia\inst{3}, 
	L.~Kaper\inst{6}, 
	L.~A.~Antonelli,
	F.~Fiore,
	A.~Gomboc,
	P.~Groot,
	F.~Hammer,
	C.~Ledoux\inst{2}, 
	E.~Maiorano,
	B.~Milvang-Jensen\inst{1}, 
	E.~Palazzi,
	E.~Pian,
	J.~Schaye,
	G.~Tagliaferri\inst{7},
	R.~A.~M.~J.~Wijers\inst{6}
}

@@

%In P84 we started the X-shooter legacy survey in GTO time. Our aim is to  build
%a sample of GRBs superior to our previous  low-resolution  survey, Fynbo et
% al.\
%2009, ApJS, 185, 526) both in terms of quantity and  quality of the spectra.
% The
%last period of GTO time on the survey was P90 (plus 6 hr of leftover GTO time
%in  P91) and for P92 a proposal like this was  accepted for continuation of the
%program in open time (concerning P93 see box 8).

@@

X-shooter is in many ways the ideal GRB follow-up instrument and indeed GRB
follow-up was one of the primary science cases behind the instrument design and
implementation. Our program secures general purpose GRB afterglow spectroscopic
follow-up that adds strong legacy value to the \textit{Swift} GRB sample. Due to
the wide wavelength range of X-shooter with the same observation cover molecular
H$_2$ absorption near the atmospheric cut-off and all the strong emission lines
from the host in the NIR arm (e.g., Friis et al., 2015, MNRAS, 451, 167). In
general, the wide wavelength coverage ensures that we always have features on
which to base a redshift measurement as long as the afterglow is brighter than
about 23 mag in either the $R$- or $z$-band. Frequently, emission lines are also
detected from the underlying host, which also provide further information such
as SFR and metallicity (the top right panel in Fig.~1 shows an example). Only
for 7 out of more than 70 secured spectra could we not measure a redshift. With
the X-shooter survey we provide {\bf metallicity measurements} for about 30\%
(Voigt-profile fits) of the $z>1.7$ events. So far we have measured
metallicities for more than 20 GRB afterglows with X-shooter. With the wide
wavelength coverage of X-shooter we can study important chemical species as
Zn\,\textsc{ii}, Cr\,\textsc{ii} and $\alpha$ elements over a much wider
redshift range than what is possible with other instruments. As an example, we
have measured a metallicity of $0.1 Z_\odot$ for GRB\,100219A at $z=4.669$
(Th{\"o}ne et al.\ 2013, MNRAS, 428, 3590), $0.02Z_\odot$ for GRB\,111008A at
$z=4.991$ (Sparre et al.\ 2014, ApJ, 785, 150) and $0.05Z_\odot$ for the
$z=5.9125$ GRB\,130606A (Hartoog et al.\ 2015, A\&A, 580, 139). Reconciling the
abundance patterns of GRB absorbers, other types of absorbers, QSO DLAs in
particular, and old stars in the Local Group is an important long-term goal (see
also Sparre et al.\ 2014, ApJ, 785, 150). Metallicities are also measured from
host emission lines (Kr{\"u}hler et al.\ 2015, A\&A, 581, A125). GRB
spectroscopy also allows us to determine the dust content of their environments,
both through analysis of the depletion pattern and through measurement of the
associated extinction (Japelj et al.\ 2015, A\&A, 451, 2050). This allows us to
quantify the dust-to-metals ratio and its evolution with redshift (e.g., De Cia
et al.\ 2013, A\&A, 560, 88; Zafar \& Watson 2013, A\&A, 560, 26). %Moreover,
% GRB
%afterglows allow investigation of random foreground absorbers, complementing
%QSO sightline studies (e.g., Schulze et
%al.\ 2012, A\&A, 546, 20).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Sample selection criteria}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Observations} \label{obs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure*}[!ht]
	\centerline{\includegraphics{figures/skymap.pdf}}
	\caption{Mollweide projection in galactic coordinates of the full sky showing
		the positions on the sky of the bursts presented in this work. The empty
		circles are the positions of all the \textit{Swift} bursts detected until the
		end of 2016. In blue is shown the position of the bursts fulfilling the sample
		criteria specified in Sect. \ref{samplecrit}. The red circles are added around
		GRBs which have been followed up by X-shooter since the commissioning of the
		instrument. Blue points with red circles represent GRBs that enter our sample
		and have X-shooter spectroscopy. The different samples are compared in Sect.
		\ref{results}. The background shows the dust maps presented in
		\citet{Schlegel1998} where the sample criteria cut with $A_V \lesssim 0.5$ mag
		is removed.}
	\label{fig:skymap}
\end{figure*}

@@

We list the overview of all the observations in Tab. \ref{tab:sample_overview}.
\input{tables/burst_overview.tex} and plot the positions off all the bursts on
the night sky in Fig. \ref{fig:skymap}. The central zone of avoidance due to
high galactic extinction cutoff is visible with an apparent isotropic
distribution of the bursts away from the galactic center except in the upper
right corner, where Earth casts a shadow as seen from the telescope.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results} \label{results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sample completeness} \label{completeness}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Using the observational characteristics of the 1244 \todo{Get number} bursts
observed by \textit{Swift} to date \cite{Gehrels2009}, and the X-ray derived
hydrogen column densities \citep{Evans2009}, we can quantify the degree to which
our sample is biased against the overall \textit{Swift} sample in terms of
intrinsic GRB properties. The values are queried from the online
\textit{Swift} database maintained\footnote{http://swift.gsfc.nasa.gov/archive/grb\_table/}.
Three main samples are of interest to compare to assess the completeness of the
follow-up campaign; the full \textit{Swift} sample consisting of all the bursts
observed by \textit{Swift} (1), the burst that fulfill the selection criteria
imposed in Sect. \ref{samplecrit} (2), and the bursts actually followed up with
X-shooter (3).

@@

For each of the samples we calculate the median and 14th and 86th percentile
which can be used as point estimates for the population distribution. These are
shown in \ref{tab:sample_properties}. It can be seen from the values that the
three samples have very similar distributions in terms of the point estimates
chosen, suggesting that our selection criteria are unbiasing against the
\textit{Swift}-sample and that additionally, the follow-up effort is unbiasing
towards the intrinsic GRB properties. This is despite spectroscopic follow-up
only being carried out in cases where either a detectable optical afterglow or a
clear counterpart are seen, which naively should be biased against dark bursts
occurring in more obscured galaxies, which is shown to exhibit different
galactic properties \citep{Perley2009, Kruhler2011, Rossi2012, Perley2013b,
	Perley2015b}.

Additionally, using a 2-sided Kolmogorov-Smirnov test (KS-test), we can assess
the degree to which the null that the two distribution are drawn from the same
distribution can be rejected. We show a graphic representation of the test
statistics in Fig. \ref{fig:p_values}. We have run the test on whether is can be
rejected that the sample criteria is drawn from the same distribution as the
\textit{Swift} sample, if the observed sample is dissimilar from the statistical
sample, and again whether the observed sample differs from the full
\textit{Swift} sample. A high p-value indicates little evidence against the null
meaning for the parameters investigated here, only the X-ray derived column
densities exhibits the highest degree of dissimilarity, but not to a significant
level. A large part of the tension between the two samples in $N_H$ is driven by
a small number of low column densities($\log(N_H) \lesssim 19 \mathrm{cm}^2$)
bursts.

@@

We therefore conclude that the sample presented here, to a high degree
represents the intrinsic properties of the GRBs in the full \textit{Swift} sample.

@@

GRB follow-up requires the immediate availability of the telescope in order to
successfully carry out spectroscopic observations. Out of the 158 \todo{Get
exact number} bursts in the statistical sample, 32($\sim$ 20 percent) \todo{Get
exact number} have not been observed due to conditions, not relating the GRB or
afterglow properties. The reasons includes unavailability of the telescope due
to technical maintenance such af mirror re-coating, a visitor rejecting the TOO
trigger, or bad weather at the observing site. Because this cut is irrespective
of the GRB properties, it is not expected that this should change sample
properties. Addressing whether this cut introduces a bias in the followed sample
is not trivial, but in terms of intrinsic GRB properties, the sample
distributions does not change by much, as shown in Sect. \ref{completeness}.
\todo{Should this be quantified somehow?}

@@

A subset of all GRBs exhibit no detectable or very faint optical afterglow \citep{Groot1998,
	Djorgovski2001, Fynbo2001}, which had been parametrized in term of "darkness"
\citep{Jakobsson2004, VanderHorst2009}. The X-ray properties of this subset has
previously been investigated \citep{DePasquale2003, Melandri2012}, finding a
slightly higher X-ray luminosity compared to the optically brights bursts
combined with similar prompt characteristics. This indicates along with
investigations of host galaxy extinction \citep{Greiner2011, Kruhler2011, Hjorth2012}, that the
extinction of the optical afterglow is primarily driven by the presence of dust,
local to the burst \todo{Local dust reference}.

@@

\begin{figure*}
	\centering
	\includegraphics[page=1, width=16cm]{figures/HI_measurements.pdf}
	\caption{Measurements of the hydrogen column-densities for all bursts with a
	clear Lyman alpha absorption system. In solid black is shown the spectrum with
	black dotted giving the corresponding 1-$\sigma$ error. Black dashed shows zero
	flux density. The solid red line is the absorption of column density equal to
	the value presented in Tab. \ref{HI_tab} with the 1-$\sigma$ interval shown
	with dashed lines.}
	\label{fig:HI1}
\end{figure*}
